---
title: Traffic Data Analysis
output: html_document
---
# Importing needed libraries

```{r}
library(rmarkdown)
library(dplyr)
```
# Loading and Processing Data

We loaded the traffic-data.csv file to the R console to read it and start discovering the basic properties

```{r}
traffic_data<- read.csv("traffic-data.csv",header=TRUE)
dim(traffic_data)
```

The traffic dataset we have got about 429982 rows and 34 variables as columns, when we look at the data we will find that the first thing to do is to eliminate the advertisement columns as it has nothing relevant to our data investigation.

We can make use of the **dplyr** library verbs to eliminate certain columns with the method select(), from the dataset we can see that we have 13 columns concerning advertisement to be eliminated.

```{r}
head(traffic_data %>% select(starts_with("ad")))
grep("ad", colnames(traffic_data))
traffic_roads<-traffic_data %>% select(-starts_with("ad"))
```
Now we can find the columns that has values with zero variance or in other words has only one unique values to remove it from our dataframe.

```{r}
sapply(traffic_roads,function(x) length(unique(x))==1)
```

We find that TRUE values reside under two columns in the dataframe which are **rd.cl** and **rd.rp.type**, so we can filter these two variables.

```{r}
var.in<-c("rd.cl","rd.rp.type")
var.out<-setdiff(names(traffic_roads),var.in)
traffic_roads<-traffic_roads[var.out]
head(traffic_roads)
```
The thing also I found it useful is to use the **strptime** function to re-formata the crawl_date variable by replacing the abbreviated day-month format with a full Y%M%D format.

```{r}
traffic_roads$crawl_date <- strptime(traffic_roads$crawl_date, format = "%a %b %e %X UTC %Y", tz = "UTC") %>% as.POSIXct()
head(traffic_roads)
```

After showing the columns I noticed that we have image columns like **rd.img** which has 300k NA values out of 400k and also we have another 2 img columns **rd.rp.img** and **rd.rp.rpImg**, we cannot make use of such columns so we included those 3 variables in the elimination process as a part of the data-cleaning.

```{r}
sum(is.na(traffic_roads$rd.img))
traffic_roads<-traffic_roads %>% select(-ends_with(".img"))
dim(traffic_roads)
head(traffic_roads)
traffic_roads<-traffic_roads %>% select(-ends_with(".rpImg"))
```

When we take a good look at how the data keeps track of the reports time, we see that we only can use **crawl_date** and **rd.rp.hr/rd.rp.mn** and none of them tells the real exact time the report took place, so thats why we needed to combine the hours and minutes together first in one column in a time format and subtract from the time exists in crawl_date we get the real report's time.

```{r}
rmins<- traffic_roads["rd.rp.mn"]
rhours<- traffic_roads["rd.rp.hr"]
hours_vector<-as.vector(rhours$rd.rp.hr)
minutes_vector<-as.vector(rmins$rd.rp.mn)
report_timestamp<- paste(hours_vector,":",minutes_vector,":00",sep="")
```

Then I used a library called **chron** to convert strings of time to real time type and format.

```{r}
library(chron)
?chron
conversion <- chron(times=report_timestamp)
traffic_roads$rp.realTime <- conversion
```

But because there are values in the **rd.rp.hr** exactly 25706 entries out of range set to <NA> thats why we removed the added column and instead we did in one step.

```{r}
traffic_roads<-traffic_roads %>% select(-starts_with("rp.realTime"))
traffic_roads <- traffic_roads %>% mutate(report_realTime = as.POSIXct(round(crawl_date - (rd.rp.hr*60*60 + rd.rp.mn*60), "mins"))) %>% select(-c(rd.rp.mn, rd.rp.hr))
```

We noticed that the column **rd.name** is formated as primary road and secondary road separated by semi-colon, thus we separated on the comma to two different columns **rd.primary_road** and **rd.secondary_road**.

Well, that lead us to a warning where there is no secondary road actually, if the report is concerning the main road only the value is set automatically to <NA>, thats why we replaced the NA to "primary" by which the entry mains that the report is sticked to the primary road.

```{r}
library("tidyr")
traffic_roads<-traffic_roads%>%separate("rd.nm",c("rd.primary_road","rd.secondary_road"),";")
traffic_roads$rd.secondary_road[is.na(traffic_roads$rd.secondary_road)] <- "primary"
head(traffic_roads)
sum(is.na(traffic_roads$rd.secondary_road))
sum((traffic_roads$rd.secondary_road)=="primary")
```

Then we need to remove the duplicate reports, which gives the same info about the same region, and that is simply by removing the duplicate values in the **rd.rp.cmid** column.

```{r}
traffic_roads<-traffic_roads[!(traffic_roads$rd.rp.cmid%>%duplicated),]
traffic_roads_unique <- traffic_roads %>% select(crawl_date, rd.ri, rd.stid) %>% unique()
glimpse(traffic_roads_unique)
```

We saw it would be more useful to separate the entries where the reporter is *bey2ollakgps* which give info about the time which the road took and its corresponding speed.

```{r}
gps_trackers<-traffic_roads[which(traffic_roads$rd.rp.nm=="bey2ollakgps"),]
head(gps_trackers)
glimpse(gps_trackers)
traffic_roads<-traffic_roads[traffic_roads$rd.rp.nm!="bey2ollakgps" , ]
sum(na.omit(traffic_roads$rd.rp.nm) == "bey2ollakgps")aggregate(x$Frequency, by=list(Category=x$Category), FUN=sum)
```

When we looked at the unique values occur in each variable we noticed that we have 3 columns which has only two distinct values which are 0 or 1, which clearly indicates a boolean variables that will be investigated later on, those columns are **rd.new**, **rd.strq** and **rd.cmrq**

```{r}
apply(traffic_roads, 2, function(x)length(unique(x)))
glimpse(traffic_roads)
```

# Trying to find some correlations

Now after we spent a long journey trying to clean the data maybe we need to get back to clean later but it is time to plot some correlation graphs

## road comments and statuses

Lets start with these columns specifically **rd.ri**, **rd.stid**, **rd.rp.stid** and **rd.rp.cmid**

```{r}
corr_vars = c("rd.ri","rd.stid","rd.rp.stid","rd.rp.cmid")
library("corrplot")
corr_numeric = cor(traffic_roads[, cor_feats], use = "complete")
corrplot(method = "shade", cor_mat)
corr_numeric
```

we find that the correlations that matters is between rd.rp.stid and rd.stid, also between rd.stid and rd.rp.cmid

## Strq and Cmrq

We can check the correlation between **rd.strq** ,**rd.cmrq** and **rd.new**.

```{r}
corr_vars = c("rd.strq","rd.cmrq","rd.new")
corr_numeric = cor(traffic_roads[, cor_feats], use = "complete")
corrplot(method = "shade", cor_mat)
corr_numeric
```
we find that the correlation between rd.strq and rd.cmrq is almost inversed, and there is a weak correlation between rd.strq and rd.new, thus rd.new is still under investigation.
